Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example

### 2023-05-24 manuscript/main.qmd
discussion part 1
still check use of "EECO" and "early Eocene"

::: {#table-1}
![](../figures/Table1.png){width="15cm"}
Table of...
:::

@@

## Improved reconstruction of temperature gradients and global average temperatures

The Bayesian model presented herein accurately reconstructs the modern, latitudinal SST gradient from a patchy and relatively small (n = 34) sample of temperature data. This is an advancement over previously used linear or quadratic approximations [e.g. @Refs], which are inherently unable to capture the sigmoidal shape of the modern SST gradient. Our generalised logistic function is flexible enough to approximate the shapes of all of those functions. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can very flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022]. However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g. @Inglis2020], and the resulting large-scale patterns are informed by the idiosyncracies of a patchy proxy record.

In our Bayesian, parametric model, informative priors on the model parameters improve the prediction of sea surface temperatures in the unsampled, very high latitudes: Notice that the upper limit of the credible interval does not increase beyond the range of the data, whereas unconstrained approaches such as splines, Gaussian processes or even standard linear regression could lead to unrealistically high upper bounds in this case [see @Rasmussen2004]. Prior information on the shape of latitudinal temperature gradients on Earth exists for all geological time periods, and should be used in any palaeoclimate reconstruction. For example, the greater amount of solar radiation per unit area in low latitudes causes Earth's latitudinal temperature gradient to be broadly negative [@Beer2008]. The ease with which such prior information can be integrated is a major advantage of our method, as the shape of the modeled gradient is controlled by four parameters which clearly relate to its magnitude, steepness and the latitude of its greatest steepness.

Palaeoclimate reconstructions are often summarised as global mean surface temperatures (GMST), providing a standardised metric for characterising the state of the Earth's climate [@Royer2004; @Inglis2020]. However, the calculation of global mean surface temperatures from sparse proxy data is susceptible to bias [@Jones2022]. By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record through inter- or extrapolation. This eliminates the problem that specific climate zones may dominate the proxy record, as intersecting the modeled temperature gradient at narrow latitudinal intervals and accounting for the varying sizes of latitudinal bands by weighting provides unbiased GMST estimates. Some deep-time GMST reconstructions account for differential sampling by calculating zonal averages, but do not use the full latitudinal range of temperatures [@Inglis2020]. We anticipate that applying this improved will significantly alter Phanerozoic, proxy-based temperature curves, particularly when applied to intervals with small or biased samples of temperature proxies. [@Royer2004; @Veizer2015; @Jones2022].

## Incorporating ecological constraints in palaeoclimate reconstructions

Our results exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021]. Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Marwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic. Our modeling framework offers a straightforward way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.

Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Fauquette2007; @Pross2012]. These methods have the problem that the thermal preferences of taxa may have changed over time. More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures. This is especially true for taxa that inhabit the warmest part of the ocean today, e.g. coral reefs \[\@\]. Although coral reefs are threatened by warming sea surface temperatures today \[\@\], it is conceivable that Eocene reef corals were adapted to a warmer climate. The fossil record indicates that reef development may have been stunted in the early Eocene, but the few early Eocene coral reefs occur in low latitudes [@Zamagni2012]. Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures (Fig. S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs. We have tried to account for that possibility by widening the temperature probability distribution for reef corals, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig. 4, Fig. S2).

This dilemma could be resolved in future studies by applying machine learning approaches to inferring quantitative temperatures from ecological and lithological proxies: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures. This approach would ultimately rely on geochemical proxies providing absolute temperature estimates, but the resulting information could be exported to other locations lacking good coverage of geochemical proxies via their fossil or sedimentary record.

<!-- This could be a neat future manuscript (although time...) so not sure if we state this idea, or just cut the previous paragraph -->

>>>>

::: {#table-1}
![](../figures/Table1.png){width="16.25cm"}
Table 1: Inferred latitudinal SST gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies. For comparison, a gradient derived from a climate model ensemble is also shown.

:::

@@

## Improved estimation of latitudinal and global palaeotemperatures

Our results show that relatively small (*n* = 10 - 20) and patchy samples of temperature data are sufficient to reconstruct greenhouse- and icehouse-type, latitudinal SST gradients, using our Bayesian model. This is an advancement over previously used linear, quadratic or Gaussian approximations [e.g. @Bijl2009; @Tierney2017], which can fit only specific types of gradients. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can very flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022]. However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g. @Inglis2020]. This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, directly generates the resulting large-scale climate patterns.

In contrast, our Bayesian, parametric model, allows for the inclusion of informative priors on the model parameters. The modeled sea surface temperature gradient thus does not strictly follow the proxy data, but instead represents a compromise between the data and our prior knowledge. In the EECO example (Fig. 4), the inclusion of informative priors improves the prediction of sea surface temperatures in the unsampled, very high latitudes: Notice that the upper limit of the credible interval does not increase beyond the range of the data, whereas unconstrained approaches such as splines, Gaussian processes or even standard linear regression could lead to unrealistically high upper bounds in this case [see @Rasmussen2004]. Prior information on the shape of latitudinal temperature gradients on Earth exists for all geological time periods. For example, the greater amount of solar radiation per unit area in low latitudes causes Earth's latitudinal temperature gradient to be broadly negative [@Beer2008]. The ease with which such prior information can be integrated is a major advantage of our method, as the shape of the modeled gradient is controlled by four parameters which clearly relate to its magnitude, steepness and the latitude of its greatest steepness.

Palaeoclimate reconstructions are often summarised as global mean surface temperatures (GMST), providing a standardised metric for characterising the state of the Earth's climate [@Royer2004; @Inglis2020]. However, the calculation of global mean surface temperatures from sparse proxy data is susceptible to bias [@Jones2022]. By modeling the temperature variation across latitudes, a complete temperature distribution along a latitudinal axis can be obtained, filling in gaps in the proxy record through inter- or extrapolation. This eliminates the common problem that specific climate zones dominate the proxy record. Reconstructing the GMST directly from the proxies would lead to an estimate biased towards the well-sampled latitudes. Calculating zonal averages alleviates this problem, but this method relies on comprehensive latitudinal coverage [@Inglis2020]. Instead, our method allows for intersecting the modeled temperature gradient at narrow latitudinal intervals, even when significant latitudinal gaps exist. Weighting the temperatures of those latitudinal intervals by area then provides unbiased GMST estimates. We anticipate that this improved method may significantly alter Phanerozoic, proxy-based temperature curves, which have often been directly calculated from the proxy record [@Royer2004; @Veizer2015]. This is particularly relevant for the early Mesozoic and older intervals, for which the spatial coverage is generally poor due to the absence of data from ocean drilling sites [@Jones2022].

## The role of ecological constraints in palaeoclimate reconstructions

Our results further exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021]. Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Markwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic. Our modeling framework offers a straightforward, efficient way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate. The model could easily be extended to include uncertainties on individual geochemical proxy data, or to differently weight different proxy records classified as more or less reliable.

Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Fauquette2007; @Pross2012]. One major limitation to these methods is that the thermal preferences of taxa may have changed over time. More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures. This is especially true for taxa that inhabit the warmest part of the ocean today, e.g. coral reefs [@kleypas1999]. Although coral reefs are threatened by warming sea surface temperatures today [@Hoegh2011], it is conceivable that Eocene reef corals were adapted to a warmer climate. The fossil record indicates that reef development may have been stunted in the early Eocene, with few early Eocene coral reefs occurring in low latitudes [@Zamagni2012]. The absence of coral reefs in higher latitudes in the early Eocene could be due to requirements in irrdiance, rather than temperature [@Muir2015]. Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig. S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs. We have tried to account for that possibility by widening the temperature probability distribution for reef corals, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig. 4, Fig. S2).

<!-- This dilemma could be resolved in future studies by applying machine learning approaches to inferring quantitative temperatures from ecological and lithological proxies: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures. This approach would ultimately rely on geochemical proxies providing absolute temperature estimates, but the resulting information could be exported to other locations lacking good coverage of geochemical proxies via their fossil or sedimentary record. -->
